COATi: statistical pairwise alignment of protein coding sequences

### 2024-01-31 alignpair_letter.tex
Update discussion.

\subsection*{Simulated Benchmark Alignments}

@@

Using $d_{seq}$ we quantified the number of perfect, imperfect and best alignments each aligner produced. We define perfect alignments as alignments with a distance of zero to their benchmark alignment ($d_{seq} = 0$) or any alignment that is evolutionary equivalent to an alignment with a distance of zero. Equivalency was included in our definition of perfect alignments to prevent the manner by which aligners break ties between multiple optimal alignments from influencing our results. Evolutionary equivalence was determined by scoring alignments using COATi's marginal model. Any alignment that had a score which matched the score of its benchmark alignment was considered perfect even if its distance to the benchmark was greater than zero.
%
Imperfect alignments are defined as alignments that are not perfect when another method successfully produces a perfect alignment for the same pair of sequences.
%
Best alignments are the alignments that the lowest distance $d_{seq}$ to the true alignment, including ties.
%
Taken together, these three statistics not only allow a direct comparison of aligners but also expose instances where all aligners fall short of achieving a perfect result.

@@

\paragraph{Empirical Data and Alignments.}

@@

The vast majority of sequence pairs (12,344) were identified as showing evidence of negative selection when they were aligned using COATi, ClustalΩ, MACSE, MAFFT, PRANK, and COATi-rev (Fig.~\ref{fig:sel-empirical}). Another 1087 sequence pairs were identified as showing evidence of positive selection when they were aligned using all 6 strategies. The most common pattern that showed disagreement between aligners were 91 sequence pairs that were identified as positively selected by COATi, PRANK, and COATi-rev. The most common singleton pattern was 64 sequence pairs identified as positively selected only by MAFFT. Notably, 36 sequence pairs were identified as positively selected and 9 sequence pairs were identified as negatively selected by by COATi and COATi-rev only.
%
In total, COATi identified 1,367 (\sim{}10\%) sequence pairs as showing evidence of positive selection. ClustalΩ identified 1,228. MACSE identified 1,183. MAFFT identified 1,340. And PRANK identified 1,349.

@@

\subsection*{Benchmark Results}

% Software comparison table
\begin{table}[!ht]
\definecolor{bestcolor}{RGB}{230,230,230}%
\centering
\caption{COATi generates better alignments than other alignment algorithms on semi-empirical benchmarks.\label{table:comp}}
%
% Copied from tables/benchmarks.tex
% footnotes and \addlinespace are manually added
\begin{tabular}{rccccc}
  \toprule
 & COATi & ClustalΩ & MACSE & MAFFT & PRANK \\ 
  \midrule
Average Alignment Error ($d_{seq}$) & \cellcolor{bestcolor}0.23\% & 2.65\% & 1.37\% & 1.67\% & 1.02\% \\
\addlinespace 
  Number of Best\textsuperscript{1} Alignments & \cellcolor{bestcolor}7050 & 3988 & 5226 & 5918 & 6128 \\ 
  Number of Perfect\textsuperscript{2} Alignments & \cellcolor{bestcolor}6795 & 3954 & 5138 & 5840 & 5860 \\ 
  Number of Imperfect\textsuperscript{3} Alignments & \cellcolor{bestcolor} 850 & 3691 & 2507 & 1805 & 1785 \\
  \addlinespace 
  RMSE for K2P Distances & \cellcolor{bestcolor}0.000694 & 0.0920 & 0.0528 & 0.0561 & 0.0305 \\ 
  Overestimated K2P Distances & \cellcolor{bestcolor}10.3\% & 52.7\% & 38.0\% & 26.2\% & 25.6\% \\ 
  \addlinespace
  F\textsubscript{1} for Negative Selection & \cellcolor{bestcolor}99.8\% & 97.6\% & 98.4\% & 98.3\% & 99.1\% \\ 
  F\textsubscript{1} for Positive Selection & \cellcolor{bestcolor}97.8\% & 77.2\% & 84.9\% & 85.5\% & 91.8\% \\ 
   \bottomrule
   \addlinespace
\multicolumn{6}{p{40em}}{\textsuperscript{1}Best alignments have the lowest $d_{seq}$ (including equivalent alignments)}\\
\multicolumn{6}{p{40em}}{\textsuperscript{2}Perfect alignments have $d_{seq} = 0$ (including equivalent alignments)}\\
\multicolumn{6}{p{40em}}{\textsuperscript{3}Imperfect alignments are alignments that are not perfect when at least one method found a perfect alignment.}\\
\end{tabular}
\end{table}

\begin{figure}[h!]
    \centering%
    \includegraphics{figures/fig-dseq-benchmarks.pdf}
    \par
    \caption{COATi's alignments were closer to the semi-empirical benchmark than other methods. A principal coordinates analysis of the average alignment distance ($d_{seq}$) between alignments generated by different methods. Not only were COATi's alignments on average closest to the benchmark, but other aligners tended to diverge from the benchmark in different directions.
%
    \label{fig:dseq-benchmarks}}
\end{figure}

@@

% NOTES:
% - Benchmark only contains seqeunces with gaps, which will make alignes look more different than they are in practice
% - Parameters and optimization
% - simulalation contained a lot of frameshifts that produed stop codons. Probably too many.
% - redo colors of figures 1 - 3.
% - define an equivalent alignment again in the benchmark results

COATi obtained better results compared to a wide variety of alignment strategies. It was significantly more accurate (lower $d_{seq}$) at inferring the empirically simulated alignments compared to other methods; all p-values were less than $1.3 \cdot 10^{-76}$ according to the one-tailed, paired Wilcoxon signed-rank tests (Supplementary Materials Figure 1). Notably, the average alignment error of the second best protocol was six times larger than COATi's. In addition, COATi produced more perfect alignments, less imperfect alignments, and more accurately inferred events of positive and negative selection (Table \ref{table:comp}). Furthermore, the estimated evolutionary divergence from the alignments retrieved by COATi was substantially less overestimated than other methods (Table \ref{table:comp}, Supplemental Materials Figure 8).

Aligning sequences using amino-acid translation via ClustalΩ obtained the highest average alignment error and also had difficulties correctly identifying positive selection and estimating evolutionary distances. These results are not surprising because alignments in amino-acid space only permit phase-3 gaps and are sensitive to frameshift artifacts, which inflate divergence estimates and impact selection detection. Conversely, we obtained better results when aligning in nucleotide space using both amino-acid aware MACSE and amino-acid agnostic MAFFT. These results indicate that aligning protein-coding sequences in nucleotide space works reasonably well if the sequences come from closely related species, such as humans and gorillas, as it supports all three gap phases and is robust to frameshift artifacts. If the sequences were further apart, we predict that amino-acid agnostic approaches would begin to produce unreasonable alignments. And finally, our PRANK protocol used a codon substitution model which should have been an improvement over both amino-acid and nucleotide models; however, PRANK's codon model only permits phase-3 gaps and is very sensitive to frameshift artifacts. In fact, PRANK refuses to align any sequence using a codon model if its length is not a multiple of three.

COATi is not a symmetric pairwise aligner, as reference sequences are more constrained than non-reference sequences. To test how well COATi performs when the roles of sequences are reversed, we realigned the 7761 benchmark sequence alignments using gorilla as the reference. Notably, COATi was only able to align 4003 sequence pairs due to the presence of early stop codons or other artifacts in 3758 simulated gorilla sequences. While the simulation algorithm prevents disrupting the reading frame and introducing frameshifts, it does not prevent early stop codons from being formed in the descendant sequence. Despite this limitation, we analyzed the 4003 alignments and compared the results across methods, including COATi using the human sequence as the reference. The results show a decrease, albeit small, in accuracy across all metrics when the low-quality sequence is used as the reference in comparison to the reverse (Supplementary Materials Tab. 6). However, the results for COATi continue to be a significant improvement over other aligners.

>>>>

\subsection*{Semi-Empirical Benchmark Alignments}

@@

Using $d_{seq}$ we quantified the number of perfect, imperfect and best alignments each aligner produced. We define perfect alignments as alignments with a distance of zero to the benchmark alignment ($d_{seq} = 0$) or any alignment that is evolutionary equivalent to an alignment with a distance of zero. Equivalency was included in our definition of perfect alignments to prevent the manner by which aligners break ties between multiple optimal alignments from influencing our results. Evolutionary equivalence was determined by scoring alignments using COATi's marginal model. Any alignment that had a score which matched the score of its benchmark alignment was considered perfect even if its distance to the benchmark was greater than zero.
%
Imperfect alignments are defined as alignments that are not perfect when another method successfully produces a perfect alignment for the same pair of sequences.
%
Best alignments are the alignments that the lowest distance $d_{seq}$ to the true alignment, including ties.
%
Taken together, these three statistics not only allow a direct comparison of aligners but also expose instances where all aligners fall short of achieving a perfect result.

@@

\subsection*{Empirical Data and Alignments}

@@

The vast majority of sequence pairs (12,344) were identified as showing evidence of negative selection when they were aligned using COATi, ClustalΩ, MACSE, MAFFT, PRANK, and COATi-rev (Fig.~\ref{fig:sel-empirical}). Another 1087 sequence pairs were identified as showing evidence of positive selection when they were aligned using all 6 strategies. The most common pattern that showed disagreement between aligners were 91 sequence pairs that were identified as positively selected by COATi, PRANK, and COATi-rev. The most common singleton pattern was 64 sequence pairs identified as positively selected only by MAFFT. Notably, 36 sequence pairs were identified as positively selected and 9 sequence pairs were identified as negatively selected by by COATi and COATi-rev only.
%
In total, COATi identified 1,367 (\sim{}10\%) sequence pairs as showing evidence of positive selection. ClustalΩ identified 1,228. MACSE identified 1,183. MAFFT identified 1,340. And PRANK identified 1,349. Note that we did not explore whether any of these inferences were biologically or statistically significant.

@@

\subsection*{Semi-Empirical Benchmark}

COATi produced better alignments than ClustalΩ, MACSE, MAFFT, and PRANK when evaluated by our benchmark data set. It produced more best alignments, more perfect alignments, and less imperfect alignments (Tab.~\ref{table:comp}).
%
COATi was significantly more accurate (lower $d_{seq}$) at inferring the benchmark alignments compared to the other methods (Tab.~\ref{table:comp}). All p-values were significant according to the one-tailed, paired Wilcoxon signed-rank tests. (P-value of COATi vs PRANK was $5.1 \times 10^{-8}$, and p-values of COATi vs the other aligners were $< 2.2 \times 10^{-16}$). Notably, the average alignment error of the second best protocol was four times larger than COATi's. We also calculated the average distance of alignments between aligners and used principle coordinate analysis (PCoA) to project the resulting distance matrix into two dimensions (Fig.~\ref{fig:dseq-benchmarks}). COATi was the closest aligner to the benchmark, and other aligners tended to diverge from the benchmark in different directions. Because benchmark alignments were derived from gap patterns estimated by different aligners, we also performed separate PCoA analyses for each source of gap patterns (Supplementary Materials, Fig.~S5). While all aligners, except ClustalΩ, do exceptionally well on their own patterns, COATi does exceptionally well on every set of patterns.

For evolutionary distances, COATi had the lowest root-mean-square error (0.000694) relative to the benchmark alignments (Tab.~\ref{table:comp}). The second lowest aligner, PRANK, had a root-mean-square error that was over 400 times larger than COATi's. COATi only overestimated  10\% of its evolutionary distances, while other aligners overestimated 26\%--53\% of their evolutionary distances.
%
Additionally, COATi more accurately inferred events of positive and negative selection (Tab.~\ref{table:comp}).

% Software comparison table
\begin{table}[!ht]
\definecolor{bestcolor}{RGB}{230,230,230}%
\centering
\caption{COATi generates better alignments than other alignment algorithms on semi-empirical benchmarks.\textsuperscript{1}\label{table:comp}}
%
% Copied from tables/benchmarks.tex
% footnotes and \addlinespace are manually added
\begin{tabular}{rccccc}
  \toprule
 & COATi & ClustalΩ & MACSE & MAFFT & PRANK \\ 
  \midrule
Average Alignment Error ($d_{seq}$) & \cellcolor{bestcolor}0.23\% & 2.65\% & 1.37\% & 1.67\% & 1.02\% \\
\addlinespace 
  Number of Best\textsuperscript{2} Alignments & \cellcolor{bestcolor}7050 & 3988 & 5226 & 5918 & 6128 \\ 
  Number of Perfect\textsuperscript{2} Alignments & \cellcolor{bestcolor}6795 & 3954 & 5138 & 5840 & 5860 \\ 
  Number of Imperfect\textsuperscript{2} Alignments & \cellcolor{bestcolor} 850 & 3691 & 2507 & 1805 & 1785 \\
  \addlinespace 
  RMSE for K2P Distances & \cellcolor{bestcolor}0.000694 & 0.0920 & 0.0528 & 0.0561 & 0.0305 \\ 
  Overestimated K2P Distances & \cellcolor{bestcolor}10.3\% & 52.7\% & 38.0\% & 26.2\% & 25.6\% \\ 
  \addlinespace
  F\textsubscript{1} Score for Negative Selection & \cellcolor{bestcolor}99.8\% & 97.6\% & 98.4\% & 98.3\% & 99.1\% \\ 
  F\textsubscript{1} Score for Positive Selection & \cellcolor{bestcolor}97.8\% & 77.2\% & 84.9\% & 85.5\% & 91.8\% \\ 
   \bottomrule
   \addlinespace
\multicolumn{6}{p{40em}}{\textsuperscript{1}Total number of sequence pairs in the benchmark was 8,261.}\\
\multicolumn{6}{p{40em}}{\textsuperscript{2}Best alignments have the lowest $d_{seq}$ (including equivalent alignments). Perfect alignments have $d_{seq} = 0$ (including equivalent alignments). Imperfect alignments are alignments that are not perfect when at least one method found a perfect alignment. }\\
\end{tabular}
\end{table}

\begin{figure}[h!]
    \centering%
    \includegraphics{figures/fig-dseq-benchmarks.pdf}
    \par
    \caption{COATi's alignments were closer to the semi-empirical benchmark than other methods. A principal coordinates analysis (PCoA) of the average alignment distance ($d_{seq}$) between alignments generated by different methods.
%
    \label{fig:dseq-benchmarks}}
\end{figure}

@@

% NOTES:
% - Benchmark only contains seqeunces with gaps, which will make alignes look more different than they are in practice
% - simulalation contained a lot of frameshifts that produed stop codons. Probably too many.
% - redo colors of figures 1 - 3.
% - define an equivalent alignment again in the benchmark results

We have developed a statistical alignment program using finite-state transducers called COATi. On empirical data, COATi produced alignments that provided reasonable biological inferences. On semi-empirical, simulated benchmarks COATi was more accurate than the other aligners/alignment strategies that we tested: ClustalΩ's amino-acid model, MACSE's amino-acid away nucleotide model, MAFFT's DNA model, and PRANK's codon model. COATi's advantage over other aligners in our tests is at least partially due to its default parameters fitting the human-gorilla coding sequences better than the other aligner's default settings. Different data sets may show different results. However, because COATi uses a statistical model with biologically meaningful parameters, users can easily optimize COATi by using parameters estimated by previous studies or estimating parameters themselves \citep[e.g.][]{zhu2022profiling}.

COATi's primary disadvantage is that its FST model was slower and consumed more resources than other methods. However, COATi's marginal model produces results similar to the FST model, was the fastest method tested, and consumed reasonable amounts of memory \citep{garcia2023dissertation}. COATi is also not a symmetric pairwise aligner, as reference sequences are more constrained than non-reference sequences.
%
In order to test whether the choice of a reference sequence impact's COATi's results, we aligned our empirical sequence pairs with COATi using gorilla as the reference. We found no biologically or statistically significant differences between the results. The most noticeable difference was that with gorilla as the reference COATi rejected more sequence pairs due to the presence of ambiguous nucleotides in gorilla sequences.

After COATi, PRANK's codon model was the second most accurate alignment method. Like COATi, PRANK also uses statistical models and supports aligning protein-coding sequences using codon models developed for phylogenetics. However, PRANK's codon model only permits phase-3 gaps and is very sensitive to frameshift artifacts. In fact, PRANK refuses to align any sequence using a codon model if its length is not a multiple of three. Similarly, ClustalΩ amino-acid model also only permits phase-3 gaps and is very sensitive to frameshift artifacts; however, it was the worst performing protocol that we tested. It had the highest average alignment error and also had difficulties correctly identifying positive selection and estimating evolutionary distances.  --- It even performed poorly on its own gap patterns. --- ClustalΩ's relatively poor performance likely derives from its default assumptions being a poor fit for our data set and from the loss of information that occurred when we translated protein-coding DNA sequences into protein sequences.

We obtained better results when aligning in nucleotide space using both amino-acid aware MACSE and amino-acid agnostic MAFFT. These results indicate that aligning protein-coding sequences in nucleotide space works reasonably well if the sequences come from closely related species, such as humans and gorillas, as it supports all three gap phases and is robust to frameshift artifacts. If the sequences were further apart, we predict that amino-acid agnostic approaches would begin to produce unreasonable alignments.

We can look at distances between alignments generated by different aligners to understand the diversity of approaches use by our alignment protocols. Aside from the cluster of different COATi methods, we detected no clusters in our analysis. Different aligners tended to spread out from one another and in different directions from the benchmark. This indicates that each protocol approached alignment inference from a different section of ``aligner space''. While each aligner, except ClustalΩ, performed well on its own gap patterns, MACSE also did well on ClustalΩ and MAFFT gap patterns, and PRANK also did well on ClustalΩ patterns. COATi did well on all patterns.

% COATi is not a symmetric pairwise aligner, as reference sequences are more constrained than non-reference sequences. To test how well COATi performs when the roles of sequences are reversed, we realigned the 7761 benchmark sequence alignments using gorilla as the reference. Notably, COATi was only able to align 4003 sequence pairs due to the presence of early stop codons or other artifacts in 3758 simulated gorilla sequences. While the simulation algorithm prevents disrupting the reading frame and introducing frameshifts, it does not prevent early stop codons from being formed in the descendant sequence. Despite this limitation, we analyzed the 4003 alignments and compared the results across methods, including COATi using the human sequence as the reference. The results show a decrease, albeit small, in accuracy across all metrics when the low-quality sequence is used as the reference in comparison to the reverse (Supplementary Materials Tab. 6). However, the results for COATi continue to be a significant improvement over other aligners.

### 2024-01-31 supplementary_materials/supplementary_materials.Rmd
Update discussion.

```{r setup, include = FALSE}
library(knitr)
library(tidyverse)
library(kableExtra)
library(here)

cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73",
    "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
set.seed(17761)

@@

```{r aln-example-a, fig.align='center', fig.cap="The FST of all possible alignments between `CTC' and `CTG'."}
knitr::include_graphics("figures/aln-example-graph.pdf")
```

```{r aln-example-b, out.width="75%", fig.align='center', fig.cap="The best alignment of `CTC' and `CTG'."}
knitr::include_graphics("figures/aln-example-path.pdf")
```

@@

tab <- tab |> pivot_longer(cols = clustalo:`coati-rev`,
    names_to="method", values_to="other") |>
    mutate(method = fct_relevel(method, "clustalo", "macse", "mafft",
        "prank", "coati-rev")) |>
    mutate(method = fct_recode(method,
        "vs ClustalΩ" = "clustalo",
        "vs MACSE" = "macse",
        "vs MAFFT" = "mafft",
        "vs PRANK" = "prank",
        "vs COATI-rev" = "coati-rev"
    ))

gg <- ggplot(tab, aes(x = coati, y = other, color = method)) + facet_wrap(vars(method))
gg <- gg + geom_point(alpha = 0.2, show.legend = FALSE)
gg <- gg + xlab("COATi") + ylab("Other")
gg <- gg + scale_color_manual(values = cbbPalette[c(2,4,6,7,8)])
gg <- gg + scale_x_continuous(trans = scales::transform_pseudo_log(0.0003, base=10),
    breaks=c(0, 0.001, 0.01, 0.1, 1), limits = c(0, 1.2))
gg <- gg + scale_y_continuous(trans = scales::transform_pseudo_log(0.0003, base=10),
    breaks=c(0, 0.001, 0.01, 0.1, 1))

@@

k2p_data1 <- k2p_data |> pivot_longer(cols = clustalo:`coati-rev`,
    names_to="method", values_to="other") |>
    mutate(method = fct_relevel(method, "clustalo", "macse", "mafft",
        "prank", "coati-rev")) |>
    mutate(method = fct_recode(method,
        "vs ClustalΩ" = "clustalo",
        "vs MACSE" = "macse",
        "vs MAFFT" = "mafft",
        "vs PRANK" = "prank",
        "vs COATI-rev" = "coati-rev"
    ))

gg <- ggplot(k2p_data1, aes(x = coati, y = other, color = method)) + facet_wrap(vars(method))
gg <- gg + geom_point(alpha = 0.2, show.legend = FALSE)
gg <- gg + xlab("COATi") + ylab("Other")
gg <- gg + scale_color_manual(values = cbbPalette[c(2,4,6,7,8)])
gg <- gg + scale_x_continuous(trans = scales::transform_pseudo_log(0.0004, base=10),
    breaks=c(0, 0.001, 0.01, 0.1, 1))
gg <- gg + scale_y_continuous(trans = scales::transform_pseudo_log(0.0004, base=10),
    breaks=c(0, 0.001, 0.01, 0.1, 1))

@@

>>>>

```{r setup, include = FALSE}
library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(ggrepel)

#cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73",
#    "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")

cbbPalette <- c(
    "COATi" = "#0077BB",
    "COATi-rev" = "#33BBEE",
    "MACSE" = "#009988",
    "MAFFT" = "#EE7733",
    "PRANK" = "#CC3311",
    "ClustalΩ" = "#EE3377",
    "Benchmark" = "#000000",
    "missing" = "#BBBBBB")

set.seed(17761)

@@

```{r aln-example-a, fig.align='center', fig.cap="The FST of all possible alignments between \"CTC\" and \"CTG\"."}
knitr::include_graphics("figures/aln-example-graph.pdf")
```

```{r aln-example-b, out.width="75%", fig.align='center', fig.cap="The best alignment of \"CTC\" and \"CTG\"."}
knitr::include_graphics("figures/aln-example-path.pdf")
```

@@

tab <- tab |> pivot_longer(cols = clustalo:`coati-rev`,
    names_to="method", values_to="other") |>
    mutate(method = fct_relevel(method, "clustalo", "macse", "mafft",
        "prank", "coati-rev")) |>
    mutate(method = fct_recode(method,
        "vs ClustalΩ" = "clustalo",
        "vs MACSE" = "macse",
        "vs MAFFT" = "mafft",
        "vs PRANK" = "prank",
        "vs COATi-rev" = "coati-rev"
    ))

pal <- cbbPalette
names(pal) <- str_c("vs ", names(pal))

gg <- ggplot(tab, aes(x = coati, y = other, color = method)) + facet_wrap(vars(method))
gg <- gg + geom_point(alpha = 0.2, show.legend = FALSE)
gg <- gg + xlab("COATi") + ylab("Other")
gg <- gg + scale_color_manual(values = pal)
gg <- gg + scale_x_continuous(trans = scales::transform_pseudo_log(0.0003, base=10),
    breaks=c(0, 0.001, 0.01, 0.1, 1), limits = c(0, 1.2))
gg <- gg + scale_y_continuous(trans = scales::transform_pseudo_log(0.0003, base=10),
    breaks=c(0, 0.001, 0.01, 0.1, 1))

@@

k2p_data1 <- k2p_data |> pivot_longer(cols = clustalo:`coati-rev`,
    names_to="method", values_to="other") |>
    mutate(method = fct_relevel(method, "clustalo", "macse", "mafft",
        "prank", "coati-rev")) |>
    mutate(method = fct_recode(method,
        "vs ClustalΩ" = "clustalo",
        "vs MACSE" = "macse",
        "vs MAFFT" = "mafft",
        "vs PRANK" = "prank",
        "vs COATi-rev" = "coati-rev"
    ))

pal <- cbbPalette
names(pal) <- str_c("vs ", names(pal))

gg <- ggplot(k2p_data1, aes(x = coati, y = other, color = method)) + facet_wrap(vars(method))
gg <- gg + geom_point(alpha = 0.2, show.legend = FALSE)
gg <- gg + xlab("COATi") + ylab("Other")
gg <- gg + scale_color_manual(values = pal)
gg <- gg + scale_x_continuous(trans = scales::transform_pseudo_log(0.0004, base=10),
    breaks=c(0, 0.001, 0.01, 0.1, 1))
gg <- gg + scale_y_continuous(trans = scales::transform_pseudo_log(0.0004, base=10),
    breaks=c(0, 0.001, 0.01, 0.1, 1))

@@

# Benchmark Results

## Alignment Distances

For each sequence pair in the benchmark, we calculated the alignment distance (d~seq~) between the benchmark alignment and the alignments generated by COATi, ClustalΩ, MACSE, MAFFT, and PRANK. We also calculated distances between all pairs of aligners. Our benchmark contained gap patterns extracted from alignments generated by different aligners. Figure \@ref(fig:bench-pcoa-split) contains the results of a metric multidimensional scaling (principle coordinate analysis; PCoA) of the matrix of average distances between aligners, separated by what type of gap patterns was used in the benchmark alignment.

Figure \@ref(fig:bench-pcoa-all) contains a principle coordinate analysis of each aligner, including different COATi models, across the entire benchmark dataset. COATi's different models produced similar alignments and cluster together along with the benchmarks.

```{r bench-pcoa-split, message = FALSE, fig.cap="COATi produced accurate alignments regardless of whether the underlying gap pattern was extracted from an alignment generated by another program. Each panel is a metric multidimensional scaling of the average distances between aligners."}
tab_dist <- read_csv(here("supplementary_materials", "data", "benchmarks-dseq-split.csv"))

methods <- unique(c(tab_dist$method1, tab_dist$method2))

tab_xy <- tab_dist |> group_by(origin) |> 
    group_modify(\(df, k) {
        m <- matrix(0, length(methods), length(methods))
        m[lower.tri(m)] <- df$dseq
        colnames(m) <- methods
        rownames(m) <- methods
        d <- as.dist(m)
        v <- cmdscale(d, 2)
        colnames(v) <- c("x", "y")
        as_tibble(v, rownames = "method")
    }) |> ungroup()

method_recode <- c(
        "Benchmark" = "benchmark",
        "COATi" = "coati-tri-mg",
        "ClustalΩ" = "clustalo",
        "MACSE" = "macse",
        "MAFFT" = "mafft",
        "PRANK" = "prank"
)

origin_recode <- c(
        "COATi Gaps" = "coati-tri-mg",
        "ClustalΩ Gaps" = "clustalo",
        "MACSE Gaps" = "macse",
        "MAFFT Gaps" = "mafft",
        "PRANK Gaps" = "prank"
)

tab <- tab_xy |> mutate(
    method = fct_recode(method, !!!method_recode),
    origin = fct_recode(origin, !!!origin_recode) )

pal <- cbbPalette

gg1 <- ggplot(tab, aes(x = x, y = y, label = method, color =  method)) +
    geom_point(show.legend = FALSE) +
    geom_text_repel(show.legend = FALSE, seed = 3)

gg1 <- gg1 + facet_wrap(vars(origin))

gg1 <- gg1 + xlab("Coordinate 1") + ylab("Coordinate 2")

gg1 <- gg1 + scale_color_manual(values = pal)

gg1 <- gg1 + coord_fixed()

gg1 <- gg1 + theme_grey(9)

gg1

```{r bench-pcoa-all, message = FALSE, fig.cap="Different COATi models produce accurate alignments that are also similar to one another."}
tab_dist <- read_csv(here("supplementary_materials", "data", "benchmarks-dseq.csv"))

methods <- unique(c(tab_dist$method1, tab_dist$method2))

m <- matrix(0, length(methods), length(methods))
m[lower.tri(m)] <- tab_dist $dseq
colnames(m) <- methods
rownames(m) <- methods
d <- as.dist(m)
v <- cmdscale(d, 2)
colnames(v) <- c("x", "y")
tab_xy <- as_tibble(v, rownames = "method")

method_recode <- c(
    "Benchmark" = "benchmark",
    "COATi-tri-mg" = "coati-tri-mg",
    "COATi-tri-ecm" = "coati-tri-ecm",
    "COATi-mar-mg" = "coati-mar-mg",
    "COATi-mar-ecm" = "coati-mar-ecm",
    "COATi-dna" = "coati-dna",
    "COATi-rev-tri-mg" = "rev-coati-tri-mg",
    "ClustalΩ" = "clustalo",
    "MACSE" = "macse",
    "MAFFT" = "mafft",
    "PRANK" = "prank"
)

tab <- tab_xy |> mutate(
    method = fct_recode(method, !!!method_recode))

pal <- c(
    "COATi-tri-mg" = "#0077BB",
    "COATi-tri-ecm" = "#0077BB",
    "COATi-mar-mg" = "#0077BB",
    "COATi-mar-ecm" = "#0077BB",
    "COATi-dna" = "#0077BB",
    "COATi-rev-tri-mg" = "#33BBEE",
    "MACSE" = "#009988",
    "MAFFT" = "#EE7733",
    "PRANK" = "#CC3311",
    "ClustalΩ" = "#EE3377",
    "Benchmark" = "#000000",
    "missing" = "#BBBBBB")

gg <- ggplot(tab, aes(x = x, y = y, label = method, color =  method)) +
    geom_point(show.legend = FALSE) +
    geom_text_repel(show.legend = FALSE, seed = 1, max.overlaps = 20, force = 2)

gg <- gg + xlab("Coordinate 1") + ylab("Coordinate 2")
gg <- gg + scale_color_manual(values = pal)
gg <- gg + coord_fixed()
gg <- gg + theme_grey(9)

### 2024-01-31 supplementary_materials/supplementary_materials.tex
Update discussion.

\caption{The FST of all possible alignments between `CTC' and `CTG'.}\label{fig:aln-example-a}
\end{figure}

@@

\caption{The best alignment of `CTC' and `CTG'.}\label{fig:aln-example-b}
\end{figure}

@@

>>>>

\caption{The FST of all possible alignments between "CTC" and "CTG".}\label{fig:aln-example-a}
\end{figure}

@@

\caption{The best alignment of "CTC" and "CTG".}\label{fig:aln-example-b}
\end{figure}

@@

\hypertarget{benchmark-results}{%
\section{Benchmark Results}\label{benchmark-results}}

\hypertarget{alignment-distances}{%
\subsection{Alignment Distances}\label{alignment-distances}}

For each sequence pair in the benchmark, we calculated the alignment distance (d\textsubscript{seq}) between the benchmark alignment and the alignments generated by COATi, ClustalΩ, MACSE, MAFFT, and PRANK. We also calculated distances between all pairs of aligners. Our benchmark contained gap patterns extracted from alignments generated by different aligners. Figure \ref{fig:bench-pcoa-split} contains the results of a metric multidimensional scaling (principle coordinate analysis; PCoA) of the matrix of average distances between aligners, separated by what type of gap patterns was used in the benchmark alignment.

\begin{figure}
\centering
\includegraphics{supplementary_materials_files/figure-latex/bench-pcoa-split-1.pdf}
\caption{\label{fig:bench-pcoa-split}COATi produced accurate alignments regardless of whether the underlying gap pattern was extracted from an alignment generated by another program. Each panel is a metric multidimensional scaling of the average distances between aligners.}
\end{figure}

Figure \ref{fig:bench-pcoa-all} contains a principle coordinate analysis of each aligner, including different COATi models, across the entire benchmark dataset. COATi's different models produced similar alignments and cluster together along with the benchmarks.

\begin{figure}
\centering
\includegraphics{supplementary_materials_files/figure-latex/bench-pcoa-all-1.pdf}
\caption{\label{fig:bench-pcoa-all}Different COATi produce accurate alignments that are also similar to one another.}
\end{figure}
